---
title: "Reproducible Research: Peer Assessment 1"
author: "Sarath"
date: "June 6, 2017"
output: html_document
---

The assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day

### Download and Extract file
```{r}
if (!file.exists("activity.csv")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","activity.zip")
  unzip(zipfile="activity.zip")
  }
```

### Loading Data to variable
```{r}
ActivityData <- read.csv("activity.csv")
```

### Load Libraries required for plotting
```{r}
library(ggplot2)
```

### Calculating Steps / Day
```{r}
StepsByDate <- tapply(ActivityData$steps, ActivityData$date, sum, na.rm=TRUE)
g <- qplot(StepsByDate, binwidth=1000, xlab="Steps/day")
print(g)
```

### Mean and Median of Steps/Day
```{r}
mean(StepsByDate, na.rm=TRUE)
median(StepsByDate, na.rm=TRUE)
```
### Calculating steps per Interval
```{r}
AvgStepsByInterval <- aggregate(x=list(steps=ActivityData$steps), by=list(interval=ActivityData$interval),
                      mean, na.rm=TRUE)
g <- ggplot(data=AvgStepsByInterval, aes(x=interval, y=steps)) +
  geom_line() +
  xlab("5-minute Interval") +
  ylab("Avg No: of Steps")
print(g)
```

### Identifying missing data
```{r}
MissingData <- is.na(ActivityData$steps)
table(MissingData)
```

### Imputing missing data
```{r}
FilledData <- ActivityData
Fillvalue <- function(steps, interval) {
  Filled <- NA
  if (!is.na(steps))
    Filled <- c(steps)
  else
    Filled <- (AvgStepsByInterval[AvgStepsByInterval$interval==interval, "steps"])
  return(Filled)
}

FilledData$steps <- mapply(Fillvalue, FilledData$steps, FilledData$interval)
```

### Steps / Day with Imputed Data
```{r}
StepsByDate <- tapply(FilledData$steps, FilledData$date, sum)
g <- qplot(StepsByDate, binwidth=1000, xlab="Steps/Day")
print(g)
```

### Recalculated Mean and Median of Steps/Day
```{r}
mean(StepsByDate)
median(StepsByDate)
```

### Finding patterns of Weekdays and Weekends
```{r}

DayType <- function(date) {
  day <- weekdays(date)
  if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
    return("weekday")
  else if (day %in% c("Saturday", "Sunday"))
    return("weekend")
  else
    stop("invalid date")
}

FilledData$date <- as.Date(FilledData$date)
FilledData$day <- sapply(FilledData$date, FUN=DayType)
```

### Plotting interval based steps on weekdays and weekends
```{r}
AvgStepsByInterval <- aggregate(steps ~ interval + day, data=FilledData, mean)
g <- ggplot(AvgStepsByInterval, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
  xlab("5-minute interval") + ylab("Number of steps")
print(g)
```